---
title: "nltt_diff functions"
author: "Richel Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nltt_diff functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In ASCII art, I show two simple LTT plots:

```{r}
# Create a simple nLTT. Datapoints created are marked 'X'
# 
#  1.0 +   X---X
#      |   |
#  0.5 X---+
#      |
#      +---+---+
#         0.5 1.0
# Create a simple nLTT. Datapoints created are marked 'X'
# 
#  1.0  +       X
#  0.75 |   X---+
#  0.5  |   |
#  0.25 X---+
#       +---+---+
#         0.5 1.0
```

The nLTT statistic (simply the difference between the two) will be around 0.25.

This vignette shows how to measure this, from:
 * two phylogenies
 * two LTT plots
 * two nLTT plots

## Two pylogenies

The two phylogenies

```{r}
#
#          +---+ a
#      +---+
#          +---+ b
#
#      +---+---+
#     -2  -1   0 time (million years)
#
#
#               
#           +---+ a
#       +---+---+ b
#           +---+ c
#               + d
#
#       +---+---+
#      -2  -1   0 time (million years)
#
```

First create those phylogenies:

```{r}
phylogeny_1 <- ape::read.tree(text = "(a:1,b:1):1;")
ape::plot.phylo(phylogeny_1, root.edge = TRUE)

phylogeny_2 <- ape::read.tree(text = "(((d:0.000000001,c:0.000000001):1,b:1):0.000000001,a:1.0):1;")
ape::plot.phylo(phylogeny_2, root.edge = TRUE)

```

Plot their nLTTs (as gray) (and the average as black):

```{r}
nLTT::nltts_plot(c(phylogeny_1, phylogeny_2), plot_nltts = TRUE)
nLTT::nltt_plot(phylogeny_1, ylim = c(0,1)); nLTT::nltt_lines(phylogeny_2)
```

From this plot it can be estimated that the area between the two gray lines is 0.25. This
means that the nLTT statistic is 0.25.

```{r}
nltt_stat <- nLTT::nltt_diff(phylogeny_1, phylogeny_2)
print(nltt_stat)

```


## Two LTT plots



## Two nLTT plots
