---
title: "nltt_diff functions"
author: "Richel Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nltt_diff functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Close to polytomies, easier to eyeball

In ASCII art, I show two simple LTT plots:

```{r}
# normalized number of lineages 
#
#  1.0 +   X---X
#      |   |
#  0.5 X---+
#      |
#      +---+---+
#         0.5 1.0   normalized time
#
#
#
# normalized number of lineages 
#
#  1.0  +       X
#  0.75 |   X---+
#  0.5  |   |
#  0.25 X---+
#       +---+---+
#         0.5 1.0   normalized time
```

The nLTT statistic (simply the difference between the two) will be around 0.25.

This vignette shows how to measure this, from:
 * two phylogenies
 * two LTT plots
 * two nLTT plots

## Two pylogenies

The two phylogenies

```{r}
#
#          +---+ a
#      +---+
#          +---+ b
#
#      +---+---+
#     -2  -1   0 time (million years)
#
#
#               
#           +---+ a
#       +---+---+ b
#           +---+ c
#               + d
#
#       +---+---+
#      -2  -1   0 time (million years)
#
```

First create those phylogenies:

```{r}
phylogeny_1 <- ape::read.tree(text = "(a:1,b:1):1;")
ape::plot.phylo(phylogeny_1, root.edge = TRUE)
phylogeny_2 <- ape::read.tree(
  text = "(((d:0.000000001,c:0.000000001):1,b:1):0.000000001,a:1.000000001):1;")
ape::plot.phylo(phylogeny_2, root.edge = TRUE)

```

Plot their nLTTs (as gray) (and the average as black):

```{r}
nLTT::nltt_plot(phylogeny_1, ylim = c(0,1)); nLTT::nltt_lines(phylogeny_2)
```

From this plot it can be estimated that the area between the two gray lines is 0.25. This
means that the nLTT statistic is 0.25.

```{r}
nltt_stat <- nLTT::nltt_diff(phylogeny_1, phylogeny_2)
nltt_stat_exact <- nLTT::nltt_diff_exact(phylogeny_1, phylogeny_2)
testthat::expect_equal(nltt_stat, 0.25, tolerance = 0.0001)
testthat::expect_equal(nltt_stat_exact, 0.25, tolerance = 0.0001)
```


## Two LTT plots

These are the original LTTs:

```{r}
# normalized number of lineages 
#
#  1.0 +   X---X
#      |   |
#  0.5 X---+
#      |
#      +---+---+
#         0.5 1.0   normalized time
#
#
#
# normalized number of lineages 
#
#  1.0  +       X
#  0.75 |   X---+
#  0.5  |   |
#  0.25 X---+
#       +---+---+
#         0.5 1.0   normalized time
```

These are the LTTs:

```{r}
# number of lineages 
#
#    2 +   X---X
#      |   |
#    1 X---+
#      |
#    0 +---+---+
#      2   1   0   time in the past
#
#
# number of lineages 
#
#    4  +       X
#    3  |   X---+
#    2  |   |
#    1  X---+
#    0  +---+---+
#       2   1   0   time in the past
```

I choose to use 'time in the past' as a time unit, as this
is what `ape` does as well.

Here we extract the branching times and number of lineages
from the phylogenies:

```{r}
b_times <- nLTT::get_branching_times(phylogeny_1)
lineages <- nLTT::get_n_lineages(phylogeny_1)
b_times2 <- nLTT::get_branching_times(phylogeny_2)
lineages2 <- nLTT::get_n_lineages(phylogeny_2)
df <- data.frame(b_times = b_times, lineages = lineages)
knitr::kable(df)
df2 <- data.frame(b_times2 = b_times2, lineages2 = lineages2)
knitr::kable(df2)
```

Here we test again if the nLTT statistic is 0.25:

```{r}
nltt_stat_exact <- nLTT::nltt_diff_exact_brts(
  b_times = b_times,
  lineages = lineages,
  b_times2 = b_times2,
  lineages2 = lineages2,
  distance_method = "abs",
  time_unit = "ago") 

testthat::expect_equal(nltt_stat_exact, 0.25, tolerance = 0.0001)
```

## Two nLTT plots

These are the original nLTTs:

```{r}
# normalized number of lineages 
#
#  1.0 +   X---X
#      |   |
#  0.5 X---+
#      |
#      +---+---+
#     0.0 0.5 1.0   normalized time
#
#
#
# normalized number of lineages 
#
#  1.00 +       X
#  0.75 |   X---+
#  0.50 |   |
#  0.25 X---+
#  0.00 +---+---+
#      0.0 0.5 1.0   normalized time
```

I choose to use 'time in the past' as a time unit, as this
is what `ape` does as well.

Here we extract the branching times and number of lineages
from the phylogenies:

```{r}
b_times <- nLTT::get_branching_times(phylogeny_1)
lineages <- nLTT::get_n_lineages(phylogeny_1)
b_times2 <- nLTT::get_branching_times(phylogeny_2)
lineages2 <- nLTT::get_n_lineages(phylogeny_2)
b_times_N <- rev(b_times) / b_times[1]
lineages_N <- lineages / tail(lineages, n = 1)
b_times2_N <- rev(b_times2) / b_times2[1]
lineages2_N <- lineages2 / tail(lineages2, n = 1)
df <- data.frame(b_times_N = b_times_N, lineages_N = lineages_N)
knitr::kable(df)
df2 <- data.frame(b_times2_N = b_times2_N, lineages2_N = lineages2_N)
knitr::kable(df2)
```

Here we test again if the nLTT statistic is 0.25:

```{r}
nltt_stat_exact <- nLTT::nltt_diff_exact_norm_brts(
  b_times_N = b_times_N,
  lineages_N = lineages_N,
  b_times2_N = b_times2_N,
  lineages2_N = lineages2_N,
  distance_method = "abs") 

testthat::expect_equal(nltt_stat_exact, 0.25, tolerance = 0.0001)
```
