---
title: "Trees with a root edge"
author: "Richel Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trees with a root edge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document intends to display a feature of the nLTT package:
if a stem is added to a phylogeny, the begin of that stem
is set as the most recent common ancestor.

Here I create a random phylogeny:

```{r}
set.seed(42)
p <- ape::rcoal(4)
```

Here is how it looks like:

```{r fig.width = 7, fig.height = 7}
ape::plot.phylo(p, root.edge = TRUE)
```

I set `root.edge` to `TRUE`, just show it is absent.

I copy that same phylogeny and add a stem:

```{r}
q <- p
q$root.edge <- 5
```

This is how that phylogeny looks like:

```{r fig.width = 7, fig.height = 7}
ape::plot.phylo(q, root.edge = TRUE)
```

When overlaying their nLTT plots, one can see
that the stem is counted:

```{r fig.width = 7, fig.height = 7}
nLTT::nltt_plot(p)
nLTT::nltt_lines(q, col = "red")
```

