---
title: "Demonstration of nLTTstat function"
author: "Richel Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Demonstration of nLTTstat function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`nLTTstat` calculates the nLTT statistic values, which is
a measure of mismatch between one phylogeny and multiple posterior
phylogenies.

In this vignette:

 * an introduction of the data we work on
 * a detailed demo how to calculate the nLTT statistic
 * a demo of how to use the function itself

## General setup

Load all libraries:

```{r}
library(nLTT)
options(warn = 2)
```

Set the random number generator seed. 

```{r}
set.seed(42)
```

Setting a random number generator seed makes the results reproducible.

A `dt` (the timestep) close to zero gives a more detailed analysis, but it will take longer.

The data we work on will be:

 * a species tree
 * a posterior, consisting of multiple phylogenies

First we construct a species tree:

```{r}
species_tree <- ape::rcoal(n = 10)
```

Then we collect four possible posterior species trees:

```{r}
posteriors <- c(
  ape::rcoal(n = 10), 
  ape::rcoal(n = 10), 
  species_tree, 
  ape::rcoal(n = 10)
)
```

Then we calculate the nLTT statistic between original species
tree and each posterior tree:

```{r}
nltt_stats <- rep(x = 0, times = length(posteriors))
for (i in 1:length(posteriors)) {
  nltt_stats[i] <- nLTT::nLTTstat(species_tree, posteriors[i])
}

inherits(species_tree, "phylo")

postr_nltts <- nLTT::get_nltt_values(phylogenies = posteriors, dt = dt)
```

One of the posterior tree is a copy of the input phylogeny.
This allows us to check our algorithm: the final nLTT statistic
should be zero for that posterior tree.


Here an nLTT plot is shown, with the original species tree in black, and
the posterior trees in other colors:

```{r fig.width = 7, fig.height = 7}
ggplot2::ggplot(
  data = stree_nltts,
  ggplot2::aes(x = stree_nltts$t, y = stree_nltts$nltt)
) + ggplot2::geom_step(
  direction = "vh",
  size = 2
) + ggplot2::geom_step(
  data = postr_nltts,
  ggplot2::aes(x = postr_nltts$t, y = postr_nltts$nltt,   colour = postr_nltts$id),
  direction = "vh",
  position = ggplot2::position_dodge(width = 0.02)
) + ggplot2::scale_x_continuous(
  limits = c(0, 1), breaks = seq(0, 1, 0.1), name = "Normalized time"
) + ggplot2::scale_y_continuous(
  limits = c(0, 1), breaks = seq(0, 1, 0.1), name = "nLTT"
) + ggplot2::ggtitle("True species tree nLTT and its posteriors")
```

One can see that one posterior tree has exactly the same nLTT plot as
the original species tree.

Of this data, the nLTT statistic is calculated twice:

 * a detailed demo how to calculate the nLTT statistic
 * a demo of how to use the `calc_nltt_stat` function

## How to calculate the nLTT statistic

The first step in calculating the nLTT statistic is to 
collect the mismatch per time point, between the one species tree 
and each of the posterior trees.
This can be done by subtracting the nLTT values of the original species tree 
from each of the posteriors:

```{r}
diff_nltts <- postr_nltts
diff_nltts$nltt <- abs(diff_nltts$nltt - rep(stree_nltts$nltt, length(posteriors)))
```

Here we plot the differences per timepoint as a boxplot:

```{r fig.width = 7, fig.height = 7}
ggplot2::ggplot(
  data = diff_nltts,
  ggplot2::aes(x = factor(diff_nltts$t), y = diff_nltts$nltt)
) + ggplot2::geom_boxplot(
) + ggplot2::scale_x_discrete(
  name = "Normalized time"
) + ggplot2::scale_y_continuous(
  limits = c(0, 1), breaks = seq(0, 1, 0.1), name = "nLTT difference"
) + ggplot2::ggtitle("Difference between true species tree nLTT and its posteriors")
```

The nLTT statistic of a posterior tree equals the surface between 
its nLTT plot and the original species tree its nLTT plot.

We can get this surface by taking the mean of all differences (technically,
this gives us the average height of a rectangle. We should also multiply by one,
which is that rectangle its width, but this is omitted 
for conciseness in the calculation):

```{r}
nltt_stats <- dplyr::summarize(dplyr::group_by(diff_nltts, id), nltt_stat = mean(nltt))
```

Here are the four nLTT statistic values as four nice-looking bar-charts:

```{r}
ggplot2::ggplot(
  data = nltt_stats, 
  ggplot2::aes(x = nltt_stats$id, y = nltt_stats$nltt_stat, fill = nltt_stats$id)
) + ggplot2::geom_bar(stat = "identity", colour = "black"
) + ggplot2::scale_y_continuous(name = "nLTT statistic"
) + ggplot2::scale_x_discrete(name = "Posterior #"
) + ggplot2::ggtitle("Posterior nLTT statistic values")
```

This plot is pretty, though not very interesting. Instead, the
distribution of the nLTT statistic gives us more useful information.
The distribution of nLTT statitic values is plotted as a boxplot here:

```{r}
ggplot2::ggplot(
  data = nltt_stats, 
  ggplot2::aes(x = 1, y = nltt_stats$nltt_stat)
) + ggplot2::geom_boxplot(
) + ggplot2::geom_point(color = nltt_stats$id
) + ggplot2::scale_y_continuous(name = "nLTT statistic"
) + ggplot2::scale_x_discrete(name = "The one posterior"
) + ggplot2::ggtitle("Posterior nLTT statistic distribution")
```

The function `calc_nltt_stats` does this in one step.

# How to use `calc_nltt_stats`

```{r}
nltt_stats <- calc_nltt_stats(
  phylogeny = species_tree, 
  others = posteriors,
  dt = dt
)
names(nltt_stats)
```

```{r}
ggplot2::ggplot(
  data = nltt_stats, 
  ggplot2::aes(x = 1, y = nltt_stats$nltt_stat)
) + ggplot2::geom_boxplot(
) + ggplot2::geom_point(color = nltt_stats$id
) + ggplot2::scale_y_continuous(name = "nLTT statistic"
) + ggplot2::scale_x_discrete(name = "The one posterior"
) + ggplot2::ggtitle("Posterior nLTT statistic distribution")
```
