---
title: "How to get the average nLTT plot"
author: "Richel Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to get the average nLTT plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Calculating the average nLTT plot of multiple phylogenies
is not a trivial tasks. The function `nltts_plot`
does exactly this.


## Examples

For all examples, the following R code must be run:

```{r}
```

### Example: Easy trees

Create two easy trees:

```{r}
newick1 <- "((A:1,B:1):2,C:3);"
newick2 <- "((A:2,B:2):1,C:3);"
phylogeny1 <- ape::read.tree(text = newick1)
phylogeny2 <- ape::read.tree(text = newick2)
```

There are very similar. `phylogeny1` has short tips:

```{r}
ape::plot.phylo(phylogeny1)
ape::add.scale.bar() #nolint
```

This can be observed in the nLTT plot:

```{r}
nLTT::nltt_plot(phylogeny1, ylim = c(0, 1))
```

As a collection of timepoints:

```{r}
nLTT::get_phylogeny_nltt_matrix(phylogeny1)
```

`phylogeny2` has longer tips:

```{r}
ape::plot.phylo(phylogeny2)
ape::add.scale.bar() #nolint
```

Also this can be observed in the nLTT plot:

```{r}
nLTT::nltt_plot(phylogeny2, ylim = c(0, 1))
```

As a collection of timepoints:

```{r}
nLTT::get_phylogeny_nltt_matrix(phylogeny2)
```


The average nLTT plot should be somewhere in the middle.

It is constructed from stretched nLTT matrices.

Here is the nLTT matrix of the first phylogeny:

```{r}
nLTT::stretch_nltt_matrix(
  nLTT::get_phylogeny_nltt_matrix(phylogeny1),
  dt = 0.20,
  step_type = "upper"
)
```

Here is the nLTT matrix of the second phylogeny:

```{r}
nLTT::stretch_nltt_matrix(
  nLTT::get_phylogeny_nltt_matrix(phylogeny2),
  dt = 0.20,
  step_type = "upper"
)
```

Here is the average nLTT matrix of both phylogenies:

```{r}
nLTT::get_average_nltt_matrix(c(phylogeny1, phylogeny2), dt = 0.20)
```

Observe how the numbers get averaged.

The same, now shown as a plot:

```{r}
nLTT::get_average_nltt(c(phylogeny1, phylogeny2), dt = 0.20, plot_nltts = TRUE)
```


### Example: Harder trees

Create two easy trees:

```{r}
newick1 <- "((A:1,B:1):1,(C:1,D:1):1);"
newick2 <- paste0("((((XD:1,ZD:1):1,CE:2):1,(FE:2,EE:2):1):4,((AE:1,BE:1):1,",
  "(WD:1,YD:1):1):5);"
)
phylogeny1 <- ape::read.tree(text = newick1)
phylogeny2 <- ape::read.tree(text = newick2)
```

There are different. `phylogeny1` is relatively simple, with two branching events happening at the same time:

```{r}
ape::plot.phylo(phylogeny1)
ape::add.scale.bar() #nolint
```

This can be observed in the nLTT plot:

```{r}
nLTT::nltt_plot(phylogeny1, ylim = c(0, 1))
```

As a collection of timepoints:

```{r}
nLTT::get_phylogeny_nltt_matrix(phylogeny2)
```

`phylogeny2` is more elaborate:

```{r}
ape::plot.phylo(phylogeny2)
ape::add.scale.bar() #nolint
```

Also this can be observed in the nLTT plot:

```{r}
nLTT::nltt_plot(phylogeny2, ylim = c(0, 1))
```

As a collection of timepoints:

```{r}
nLTT::get_phylogeny_nltt_matrix(phylogeny2)
```


The average nLTT plot should be somewhere in the middle.

It is constructed from stretched nLTT matrices.

Here is the nLTT matrix of the first phylogeny:

```{r}
nLTT::stretch_nltt_matrix(
  nLTT::get_phylogeny_nltt_matrix(phylogeny1),
  dt = 0.20,
  step_type = "upper"
)
```

Here is the nLTT matrix of the second phylogeny:

```{r}
nLTT::stretch_nltt_matrix(
  nLTT::get_phylogeny_nltt_matrix(phylogeny2),
  dt = 0.20,
  step_type = "upper"
)
```

Here is the average nLTT matrix of both phylogenies:

```{r}
nLTT::get_average_nltt_matrix(c(phylogeny1, phylogeny2), dt = 0.20)
```

Observe how the numbers get averaged.

The same, now shown as a plot:

```{r}
nLTT::get_average_nltt(
  c(phylogeny1, phylogeny2),
  dt = 0.20,
  plot_nltts = TRUE
)
```
